Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-07-27 threshold/draft.Rnw
New intro

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{
J. Baglama\\
Department of Mathematics,\\University of Rhode Island
\and
B. Lewis\\
Paradigm4, Inc.\\
\and
A. Poliakov\\
Paradigm4, Inc.
}

@@

%%<<eval=TRUE, echo=FALSE>>=
%%Sys.setenv(TEXINPUTS=getwd(),
%%           BIBINPUTS=getwd(),
%%           BSTINPUTS=getwd())
%%@

@@

Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance, and other fields
can produce correlation matrices too large to explicitly form and tractably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.

@@

Finding highly-correlated pairs among a large set of vectors is an important
part of many applications. For instance, subsets of highly-correlated vectors
of gene expression values may be used in the discovery networks of genes
relevant to particular biological processes\cite{genomics}. Identification of
highly-correlated pairs may be used in feature selection algorithms for machine
learning applications\cite{ml1, ml2}. The problem is also important to time
series and image processing applications, see for example\cite{timeseries,
svd-similarity}.

@@

Note that each term in the sum is nonnegative and the weights $\sigma_j^2$,
$j=1,2,\ldots,k$ are nonincreasing and defined by the covariance structure of
the data.  Given a correlation threshold $t$, $0<t<1$, Equations \ref{cordist}
and \ref{vdist} together suggest that pairs of vectors can be excluded from
consideration whenever their projected distance is too large.  For example if
$\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 \ge 2(1-t)$, then we can conclude from
just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
correlation threshold $t$.

In practice, many pairs of vectors may be ruled out in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.  Approaches that
cheaply omit pairs of vectors from consideration are called \emph{pruning
methods}, and a variety of pruning methods have been described elsewhere for
the thresholded correlation problem; see, for
example\cite{prune1, prune2, prune3}
and the references therein.  The approach described here prunes along an
intuitive path of decreasing variance defined by the data and computed by the
SVD.

>>>>

\title{Efficient Thresholded Correlation using Truncated Singular Value Decomposition}
\author{
James Baglama\\
Department of Mathematics,\\University of Rhode Island
\and
Bryan Lewis\\
Paradigm4, Inc.\\
\and
Alex Poliakov\\
Paradigm4, Inc.
}

@@

@@

Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, financial and other time
series and other applications can produce correlation matrices
too large to explicitly form and tractably compute. Often, only values
corresponding to highly-correlated vectors are of interest, and those values
typically make up a small fraction of the overall correlation matrix. We
present a method based on the singular value decomposition (SVD) and its
relationship to the data covariance structure that can efficiently compute
thresholded subsets of very large correlation matrices.

@@

Finding highly-correlated pairs among a large set of vectors is an important
part of many applications. For instance, subsets of highly-correlated vectors
of gene expression values may be used in the discovery of networks of genes
relevant to particular biological processes\cite{genomics}.
Identification of highly-correlated pairs may be used in feature selection
algorithms for machine learning applications\cite{ml1, ml2}. It's also
important to many time series and image processing applications, see for
example\cite{timeseries, svd-similarity} and references therein.

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasable for high-dimensional data arising in modern genomics and other
applications. Considerable attention has been devoted to methods called
\emph{pruning methods} that cheaply prune away all but the most
highly-correlated pairs of vectors.  A variety of pruning methods have been
described for the thresholded correlation problem; see, for
example\cite{prune1, prune2} and the references therein. Related methods
approximate a set containing the most highly-correlated pairs\cite{prune3,
timeseries}. A method of Wu, et al.\cite{svd-similarity}, conceptually very
similar to our approach, uses the singular value decomposition (SVD) to find
pairs of close vectors with respect to a distance metric (instead of
correlation).  Our method described below prunes along an intuitive path of
decreasing variance defined by the data and computed by the SVD.

@@

Note that each term in the sum is nonnegative and the weights $\sigma_j^2$,
$j=1,2,\ldots,k$ are nonincreasing and defined by the covariance structure of
the data.  Given a correlation threshold $t$, $0<t<1$, Equations \ref{cordist}
and \ref{vdist} together suggest that pairs of vectors can be excluded from
consideration whenever their projected distance is too large.  For example if
$\sigma_1^2 |e_i^Tv_{1} - e_j^Tv_{1}|^2 \ge 2(1-t)$, then we can conclude from
just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
correlation threshold $t$.
In practice, many pairs of vectors may be pruned in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.
